PLAE enables quick and easy searching across one million ocular transcriptomes

### 2022-03-28 manuscript_00.Rmd
text updates

## New studies identified since 2019 data run

We added X studies from X species and X tissues. 
## scEiaD re-created and pipeline updated

The scEiaD pipeline is largely the same as in Swamy et al. [@swamy_building_2021]. Briefly, we collect the raw fastq files and quantify with the kallisto / bustools system [@melsted_barcode_2019; @melsted_modular_2019]. The quantified counts are merged into one matrix that is used for downstream processing with scvitools to batch correct the data and Seurat for quality control, clustering, and 2D UMAP visualization. We use our previous xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamy_building_2021].

The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300 Second, we use the DecontX algorithm [@yang_decontamination_2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination. Third, we altered the procedure for automatically aligning gene names across species (see below for details). Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process.

Two custom cutoffs for number of unique genes identified were used for the E12 chicken data and the cornea dataset (SRP362101) as an abnormally high number of cells were returned from our pipeline. We set the cutoffs for these studies to be 1000 and 800, respectively, as these values resulted in approximately the same number of cells being returned as the authors reported in their papers.

@@

As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify, we must align shared genes. We found homologs/orthologs by pulling from BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference. We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus Musculus*. This generates a table linked by Ensembl gene IDs and gene names. We then use a full join on non-duplicated Ensembl IDs between mouse and human. As we noticed a small number of gene names failed to be linked in this manner we ran another join on the remaining non-aligned genes, using gene name. This procedure gave us 17,769 shared human and mouse genes. To align the other species we again joined on Ensembl gene ID, removing genes where one chicken or macaque gene aligned to multiple human/mouse genes. In cases where multiple chicken or macaque genes aligned to one human/mouse gene, then we aggregated the counts by sum. In cases where there was no corresponding chicken or macaque gene, we filled in zeros. After these steps we had, as expected, 17,769 genes.

@@

Every study brought into scEiaD was hand curated to label biological sample, published paper ID (where available), organ, tissue (e.g. Cornea), source (iPSC based or tissue), single cell platform (e.g. 10Xv1 or DropSeq). We also, where possible, annotated retina region (e.g. fovea), sex, and age or developmental stage. To bring across individual cell type labels (e.g. Rod, Muller Glia), we wrote custom code for each study that made this table available to link to the individual cells. We curated two types of cell type labels: "CellType" and "PubCellType" where the former is our normalized cell types taken from the published labels. In this field we normalized names, for example changin "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level). We did retain the original published (except for the broader name normalization) labels under PubCellType.

## Pan mouse and human non-ocular references

The Tabula Muris and

@@

@@

## Updates to the scEiaD backend
Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types. We updated the scEiaD resource for the PLAE v0.90 web app in six important ways. First, we added two chicken retina (*Gallus gallus*) studies. Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX). Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 200 to 300, which reduces the number of cells that pass QC but improves the quality. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.

After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).

@@

There are a few other differences in between PLAE and Spectacle. First, there are no non-ocular datasets in Spectacle. Second, the cell types labels have not been harmonized. Third, as this resource uses Seurat objects, the app takes many seconds to load an individual dataset and many more to display the data.

@@

## Rich web-available visualizations

PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values. So thus a user can, on the left, see Crx expression across over a million cells, and on the right, see cell type labels for those same million cells. The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells. Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area.

For users interested in seeing how expression of a gene relates to study or other covariates, the "Exp Plot" view displays expression by cell type, predicted cell type, or cluster. The values can then be split and colored by multiple covariates like study or organism. This allows a user to see whether a gene pattern is consistent across studies or species. We also offer a "Dot Plot" view which can display a large number of genes in a space efficient manner. This allows a users to see whether a set of genes has related (or not) expression patterns across multiple cell types.

While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina.

>>>>

## scEiaD re-built and pipeline updated

The scEiaD pipeline is largely the same as used in Swamy et al. [@swamy_building_2021]. Briefly, we collect the raw fastq files and quantify with the kallisto / bustools system [@melsted_barcode_2019; @melsted_modular_2019]. The quantified counts are merged into one matrix that is used for downstream processing with scvitools to batch correct the data and Seurat for quality control, clustering, and 2D UMAP visualization. We use our previous xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamy_building_2021].

The following alterations were made compared to our previous publication. First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300. Second, we use the DecontX algorithm [@yang_decontamination_2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination. This is run across each study. Third, we altered the procedure for aligning gene names across species (see below for details). Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process.

Two custom cutoffs for number of unique genes identified were used for E12 chicken data (SRP286543) and a cornea dataset (SRP362101) as an abnormally high number of cells were returned from our pipeline. We set the cutoffs for these studies to be 1000 and 800, respectively, as these values resulted in approximately the same number of cells being returned as the authors reported in their papers.

@@

We found homologs/orthologs by pulling from BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference. We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus Musculus*. This generates a table linked by Ensembl gene IDs (e.g. ENSG00000105392) and gene names (e.g. CRX). We then use a full join on non-duplicated Ensembl IDs between mouse and human. As we noticed a small number of gene names failed to be linked in this manner we ran another join on the remaining non-aligned genes, using gene name. This procedure gave us 17,769 shared human and mouse genes. To align the other species we again joined on Ensembl gene ID, removing genes where one chicken or macaque gene aligned to multiple human/mouse genes. In cases where multiple chicken or macaque genes aligned to one human/mouse gene, then we aggregated the counts by sum. In cases where there was no corresponding chicken or macaque gene, we filled in zeros. After these steps we had, as expected, 17,769 genes.

@@

Every study brought into scEiaD was hand curated to label biological sample, published paper ID (where available), organ, tissue (e.g. Cornea), source (iPSC based or tissue), single cell platform (e.g. 10Xv1 or DropSeq). We also, where possible, annotated retina region (e.g. fovea), sex, and age or developmental stage. To bring across individual cell type labels (e.g. Rod, Muller Glia), we wrote custom code for each study that made this table available to link to the individual cells. We curated two types of cell type labels: "CellType" and "PubCellType". In the "CellType" field we normalized names, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level). The original published (except for the broader name normalization) labels are retained in PubCellType.

## Differential expression testing

@@

Similar to our related bulk RNA-seq resource, eyeIntegration, PLAE is built with the R-based Shiny platform. It is hosted on a cloud server instance. We make all data files and code available (www.github.com/plaeApp) to enable local install of the resource, if desired.

@@

## Updates to the scEiaD database
Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types. We updated the scEiaD resource for the PLAE v0.90 web app in six important ways. First, we added two chicken retina (*Gallus gallus*) studies. Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX). Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 200 to 300, which reduces the number of cells that pass QC but improves the quality. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.

After the dataset and quality control updates the scEiaD 2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).

@@

There are a few other differences in between PLAE and Spectacle. First, there are no non-ocular datasets in Spectacle. Second, the cell types labels have not been harmonized. Third, as this resource uses Seurat objects, the app takes many seconds to load an individual dataset and many more to display the data. Fourth, extensive manual curation was done to normalize cell type names, developmental stage, and other important experimental details (see Methods).

@@

## Rich interactive visualizations

PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell value plots. For example, a user can see Crx expression across over a million cells on the left and cell type labels for those same million cells on the right. The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells. Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area.

For users interested in seeing how expression of a gene relates to study or other covariates, the "Exp Plot" (short for Expression) view displays expression by cell type, predicted cell type, or cluster. The values can then be split and colored by multiple covariates like study or organism. This allows a user to see whether a gene pattern is consistent across studies or species. We also offer a "Dot Plot" view which can display expression as dots by gene and user-selected covariate (the dot color is level of expression and size is percentage). This allows a users to see whether a semi-large of genes has related (or not) expression patterns across multiple cell types.

While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer. The major cells of the retina through to the RPE are displayed in a drawn cross-section. Each cell type is colored by the amount of expression of a user given gene.
